A Multi-Scale Analysis of 27,000 Urban Street Networks

### 2017-12-16 draft/v2/latex/paper.tex
update street network measures descriptions

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. \emph{Average street length}, the mean edge length in the undirected representation of the graph, serves as a linear proxy for block size and indicates how fine-grained or coarse-grained the network is. \emph{Node density} is the number of nodes divided by the area covered by the network. \emph{Intersection density} is the density of the set of nodes with more than one street emanating from them (thus excluding dead-ends). \emph{Edge density} is the sum of all edge lengths divided by area, and the physical \emph{street density} is the sum of all edges in the undirected representation of the graph divided by area. These density measures suggest how fine-grained the network is. Finally, \emph{average circuity} represents the average ratio between each edge length and the straight-line distance between the two nodes it links (cf. [redacted]).

The \emph{average node degree}, or mean number of edges incident to each node, quantifies how well the nodes are connected. Similarly, but more concretely, the \emph{average streets per node} measures the mean number of physical streets (i.e., edges in the undirected representation of the graph) that emanate from each intersection and dead-end. This adapts the average node degree for physical form rather than directed flow. The statistical and spatial distributions of number of streets per node characterize the type, prevalence, and dispersion of intersection connectedness in the network. \emph{Connectivity} measures the minimum number of nodes or edges that must be removed from a connected network to disconnect it. The \emph{average node connectivity} (ANC) of a network---the mean number of internally node-disjoint paths between each pair of nodes---represents the expected number of nodes that must be removed to disconnect a randomly selected pair of non-adjacent nodes \cite{beineke_average_2002,dankelmann_bounds_2003}.

The \emph{clustering coefficient} of a node is the ratio of the number of edges between its neighbors to the maximum possible number of edges that could exist between these neighbors \cite{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this ratio by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the network's nodes. \emph{Betweenness centrality} indicates the importance of nodes in a network by evaluating the number of shortest paths that pass through each node \cite{barthelemy_betweenness_2004,huang_trajgraph:_2016, zhong_revealing_2017}. The maximum betweenness centrality (MBC) in a network specifies the proportion of shortest paths that pass through the most important node. This is an indicator of resilience: networks with a high MBC are more prone to failure or inefficiency should this single choke point fail. Finally, \emph{PageRank} ranks nodes based on the structure of incoming links and the rank of the source node \cite{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

New York             & 8937          & 1.06         & 0.04                       & 0.18           & 0.62        & 0.2         & 34.44                 & 8.84                  & 148                   & 2.86             \\
Atlanta              & 6850          & 1.1          & 0.04                       & 0.32           & 0.58        & 0.09        & 18.39                 & 6.16                  & 186                   & 2.45             \\

@@

Philadelphia         & 5132          & 1.08         & 0.05                       & 0.17           & 0.63        & 0.2         & 26.65                 & 7.3                   & 159                   & 2.87             \\
Boston               & 4852          & 1.09         & 0.05                       & 0.2            & 0.68        & 0.11        & 24.23                 & 6.44                  & 154                   & 2.71             \\

@@

Houston              & 4303          & 1.08         & 0.04                       & 0.2            & 0.57        & 0.22        & 33.49                 & 8.62                  & 145                   & 2.83             \\
Detroit              & 3461          & 1.07         & 0.04                       & 0.15           & 0.63        & 0.22        & 31.1                  & 8.56                  & 159                   & 2.93             \\

@@

Miami                & 3204          & 1.1          & 0.05                       & 0.17           & 0.59        & 0.23        & 40.54                 & 10.61                 & 149                   & 2.89             \\
Phoenix              & 2968          & 1.09         & 0.05                       & 0.2            & 0.62        & 0.17        & 35.31                 & 9.1                   & 150                   & 2.77             \\

@@

Seattle              & 2617          & 1.07         & 0.03                       & 0.3            & 0.54        & 0.16        & 31.57                 & 8.2                   & 143                   & 2.57             \\
Tampa                & 2479          & 1.1          & 0.05                       & 0.2            & 0.58        & 0.21        & 31.35                 & 8.46                  & 153                   & 2.83             \\
St. Louis            & 2392          & 1.1          & 0.04                       & 0.22           & 0.62        & 0.15        & 29.68                 & 8.16                  & 154                   & 2.73             \\
Pittsburgh           & 2345          & 1.09         & 0.04                       & 0.23           & 0.6         & 0.16        & 23.57                 & 6.71                  & 165                   & 2.72             \\

@@

Cincinnati           & 2040          & 1.07         & 0.03                       & 0.31           & 0.54        & 0.14        & 17.96                 & 6.1                   & 186                   & 2.51             \\

@@

Charlotte            & 1920          & 1.08         & 0.04                       & 0.3            & 0.57        & 0.11        & 21                    & 6.43                  & 170                   & 2.51             \\

@@

Indianapolis         & 1828          & 1.08         & 0.05                       & 0.23           & 0.59        & 0.17        & 27.62                 & 7.63                  & 157                   & 2.7              \\
Kansas City          & 1756          & 1.06         & 0.04                       & 0.21           & 0.58        & 0.2         & 32.09                 & 8.57                  & 152                   & 2.79             \\
Denver               & 1729          & 1.07         & 0.05                       & 0.2            & 0.57        & 0.22        & 40.6                  & 9.84                  & 138                   & 2.84             \\
Orlando              & 1548          & 1.11         & 0.06                       & 0.2            & 0.61        & 0.18        & 26.3                  & 7.44                  & 163                   & 2.79             \\
San Antonio          & 1547          & 1.07         & 0.05                       & 0.17           & 0.6         & 0.21        & 28.33                 & 7.91                  & 162                   & 2.87             \\
Nashville            & 1460          & 1.08         & 0.03                       & 0.27           & 0.59        & 0.14        & 19.08                 & 6.1                   & 181                   & 2.6              \\
Milwaukee            & 1413          & 1.06         & 0.06                       & 0.14           & 0.55        & 0.3         & 28.27                 & 7.81                  & 157                   & 3.03 \\

>>>>

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. The \emph{average street segment length} is a linear proxy for block size and specifies the network's grain. \emph{Node density} divides the node count by the network's area, while \emph{intersection density} excludes dead-ends to represent the density of street junctions. \emph{Edge density} divides the total directed network length by area, while \emph{street density} does the same for an undirected representation of the graph (to not double-count bidirectional streets). \emph{Average circuity} measures the ratio of edge lengths to the great-circle distances between the nodes these edges connect, indicating the street pattern's curvilinearity (cf. [redacted]).

The network's \emph{average node degree} quantifies connectedness in terms of the average number of edges incident to its nodes. The \emph{average streets per node} adapts this for physical form rather than directed circulation. It measures the average number of physical streets that emanate from each node (i.e., intersection or dead-end). The distribution and proportion of streets per node characterize the type, pervasiveness, and spatial dispersal of network connectedness and dead-ends. Connectivity represents the fewest number of nodes or edges that will disconnect the network if they are removed and is thus an indicator of resilience. A network's \emph{average node connectivity} (ANC)---the mean number of internally node-disjoint paths between each pair of nodes---more usefully represents how many nodes must be removed on average to disconnect a randomly selected pair of nodes \cite{beineke_average_2002,dankelmann_bounds_2003}. Brittle points of vulnerability characterize networks with low average connectivity.

A node's \emph{clustering coefficient} represents the ratio between its neighbors' links and the maximum number of links that could exist between them \cite{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the nodes. \emph{Betweenness centrality} evaluates how many of the network's shortest paths pass through some node (or edge) to indicate its importance \cite{barthelemy_betweenness_2004,huang_trajgraph:_2016, zhong_revealing_2017}. A network's \emph{maximum betweenness centrality} (MBC) measures the share of shortest paths that pass through the network's most important node: higher maximum betweenness centralities suggest networks more prone to inefficiency if this important choke point should fail. Finally, \emph{PageRank} ranks nodes based on the structure of incoming links and the rank of the source node \cite{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

New York             & 8937          & 1.06         & 0.04                       & 0.18           & 0.62        & 0.20        & 34.44                 & 8.84                  & 148                   & 2.86             \\
Atlanta              & 6850          & 1.10         & 0.04                       & 0.32           & 0.58        & 0.09        & 18.39                 & 6.16                  & 186                   & 2.45             \\

@@

Philadelphia         & 5132          & 1.08         & 0.05                       & 0.17           & 0.63        & 0.20        & 26.65                 & 7.30                  & 159                   & 2.87             \\
Boston               & 4852          & 1.09         & 0.05                       & 0.20           & 0.68        & 0.11        & 24.23                 & 6.44                  & 154                   & 2.71             \\

@@

Houston              & 4303          & 1.08         & 0.04                       & 0.20           & 0.57        & 0.22        & 33.49                 & 8.62                  & 145                   & 2.83             \\
Detroit              & 3461          & 1.07         & 0.04                       & 0.15           & 0.63        & 0.22        & 31.10                 & 8.56                  & 159                   & 2.93             \\

@@

Miami                & 3204          & 1.10         & 0.05                       & 0.17           & 0.59        & 0.23        & 40.54                 & 10.61                 & 149                   & 2.89             \\
Phoenix              & 2968          & 1.09         & 0.05                       & 0.20           & 0.62        & 0.17        & 35.31                 & 9.10                  & 150                   & 2.77             \\

@@

Seattle              & 2617          & 1.07         & 0.03                       & 0.30           & 0.54        & 0.16        & 31.57                 & 8.20                  & 143                   & 2.57             \\
Tampa                & 2479          & 1.10         & 0.05                       & 0.20           & 0.58        & 0.21        & 31.35                 & 8.46                  & 153                   & 2.83             \\
St. Louis            & 2392          & 1.10         & 0.04                       & 0.22           & 0.62        & 0.15        & 29.68                 & 8.16                  & 154                   & 2.73             \\
Pittsburgh           & 2345          & 1.09         & 0.04                       & 0.23           & 0.60        & 0.16        & 23.57                 & 6.71                  & 165                   & 2.72             \\

@@

Cincinnati           & 2040          & 1.07         & 0.03                       & 0.31           & 0.54        & 0.14        & 17.96                 & 6.10                  & 186                   & 2.51             \\

@@

Charlotte            & 1920          & 1.08         & 0.04                       & 0.30           & 0.57        & 0.11        & 21.00                 & 6.43                  & 170                   & 2.51             \\

@@

Indianapolis         & 1828          & 1.08         & 0.05                       & 0.23           & 0.59        & 0.17        & 27.62                 & 7.63                  & 157                   & 2.70             \\
Kansas City          & 1756          & 1.06         & 0.04                       & 0.21           & 0.58        & 0.20        & 32.09                 & 8.57                  & 152                   & 2.79             \\
Denver               & 1729          & 1.07         & 0.05                       & 0.20           & 0.57        & 0.22        & 40.60                 & 9.84                  & 138                   & 2.84             \\
Orlando              & 1548          & 1.11         & 0.06                       & 0.20           & 0.61        & 0.18        & 26.30                 & 7.44                  & 163                   & 2.79             \\
San Antonio          & 1547          & 1.07         & 0.05                       & 0.17           & 0.60        & 0.21        & 28.33                 & 7.91                  & 162                   & 2.87             \\
Nashville            & 1460          & 1.08         & 0.03                       & 0.27           & 0.59        & 0.14        & 19.08                 & 6.10                  & 181                   & 2.60             \\
Milwaukee            & 1413          & 1.06         & 0.06                       & 0.14           & 0.55        & 0.30        & 28.27                 & 7.81                  & 157                   & 3.03 \\
